---
title: "EDS 213 Group Project"
author: "Halina Do-Linh, Marie Rivers, Allie Cole"
date: "10/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(httr)
library(dataRetrieval)
library(metajam)
library(lubridate) #to change date types
library(utils) #to save our df as a csv
```


### Pulling Down Data with Metajam
- Each data set is loaded into the `data` folder and then a corresponding folder specifically for that data set 

#### Otter Data  
```{r, eval = FALSE}
# Santa Barbara Coastal LTER, Daniel C Reed, Shannon Harrer, Clint J Nelson, and Robert J Miller. 2021. SBC LTER: Reef: Sightings of sea otters (Enhydra lutris) near Santa Barbara and Channel Islands, ongoing since 2007.
# data url for sea otter data

data_lter_url <- "https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-sbc%2F61%2F8%2F85e13b27599bf1ed68c7930674e1e7bd"

metajam::download_d1_data(data_lter_url, "data", "counts_sb")
```
```{r}
otter_folder <- file.path("data", "counts_sb")
```



#### Urchin Data
```{r, eval=FALSE}
#Santa Barbara Coastal LTER, Steven C Schroeter, John Douglas Dixon, Thomas Ebert, and John Richards. 2021. SBC LTER: Settlement of urchins and other invertebrates, ongoing since 1990. LTER Network Member
# data url for urchin data

data_urchins_url <- "https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-sbc%2F52%2F10%2F6f1d97294f041ee3d39cda102e0223af"

metajam::download_d1_data(data_urchins_url, "data", "urchins")
```

```{r}
urchin_folder <- system.file("data", "counts_sb")
```



### Visualizing and Cleaning the Data
  
#### Otter Data
```{r}
# read in and visualize otter data
# clean data to only include interested years (2007-2011) and attributes (year, site, species, and counts)
otter_LTER <- read_d1_files(otter_folder, fnc = "read_csv")
otter_LTER_data <- otter_LTER$data


otter_LTER_data <- otter_LTER_data %>%
  filter(YEAR <= 2011) %>%
  filter(COUNT > 0) %>% 
  mutate(SITE = SBC_SITE) %>%
  mutate(SPECIES = "sea otter") %>%
  select(DATE, SITE, SPECIES, COUNT)

ggplot(data = otter_LTER_data, aes(x = DATE, y = COUNT)) +
  geom_point() +
  labs(title = "Otter Counts in the Santa Barbara Channel", 
       x = "Date", 
       y = "Number of Otters seen")

ggsave(here("figs", "otter_counts.png"))
```

#### Urchin Data
```{r}
# read in and visualize urchin data
# clean data to format dates as YYYY-MM-DD, only include interested years (2007-2011) and attributes (year, site, species, and counts)
urchins <- read_csv("https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-sbc%2F52%2F10%2F6f1d97294f041ee3d39cda102e0223af", na = "-99999") %>%
  mutate(DATE = lubridate::mdy(DATE_RETRIEVED)) %>%
  filter(DATE >= "2007-01-01") %>%
  filter(TOTAL_URCHINS > 0) %>% 
  mutate(SPECIES = "urchin") %>%
  mutate(COUNT = TOTAL_URCHINS) %>%
  select(DATE, SITE, SPECIES, COUNT)
 
urchins
ggplot(data = urchins, aes(x = DATE, y = COUNT)) +
  geom_point() +
   labs(title = "Urchin Counts in the Santa Barbara Channel", 
       x = "Date", 
       y = "Number of Urchin seen")

ggsave(here("figs", "urchin_counts.png"))
```


### Combining Data Sets into one Data Frame 

```{r}
# combine the cleaned otter and urchin data sets
combined_data <- rbind(urchins, otter_LTER_data)

ggplot(data = combined_data, aes(x = DATE, y = COUNT, color = SPECIES)) +
  geom_point() +
   labs(title = "Urchin and Otter Counts in the Santa Barbara Channel", 
       x = "Date", 
       y = "Species Count", 
       color = "Species")

ggsave(here("figs", "both_counts.png"))
```

### Saving the df as a .csv file 

```{r}
# write newly created data set to a .csv file, so that we can add this to our github 
write.csv(combined_data,"clean_data\\sb_otter_urchin_counts.csv", row.names = TRUE)
```

### Final Thoughts

- We did not have time to do a more in depth analysis, as the time was short and we spent a long time looking for data. If we had been able to access the USGS data set, we think that this could have been a much more comprehensive review of the two species populations. 
- Further, if we had been able to use the USGS data we would have expanded our sites in include more of the West Coast, not just the Santa Barbara Channel. We found a few data sets that ranged from Canada down to Mexico, however with the short time span we did not request access from USGS. 










